---
title: "Data Scraping"
author: "Linus Jen"
date: "June 23, 2020"
output: html_document
---

```{r packages, message = F, warning = F}
library(httr)
library(lubridate)
library(rvest)
library(stringr)
#install.packages("data.table")
library(data.table)
library(ggplot2)
library(dplyr)
library(knitr)
```

Pulling the 2019 schedule - for practice.

```{r r/g}
# Create storage for data frame
full_schedule_2019 = list()

# Set the main page to navigate from
full_teams_page = read_html("https://www.baseball-reference.com/teams/")

# List of all teams
teams = full_teams_page %>% html_nodes("#teams_active .left a") %>% html_text

# Moving from team page to the full schedule page

# Set starting point
teampage = html_session("http://www.baseball-reference.com/teams/")

# Create a loop for each team
for(i in teams) {
  # First, save each team's schedule
  team_sched = teampage %>% follow_link(i) %>% follow_link("2019") %>% follow_link("Schedule and Results") %>% read_html()
  
  # Then, save the important parts
  full_schedule_2019[[i]]$game_num = team_sched %>% html_nodes("th.right") %>% html_text %>% as.numeric()
  full_schedule_2019[[i]]$date = team_sched %>% html_nodes("th+ .left a") %>% html_text %>% paste0(", 2020") %>% as.Date(format = "%A, %b %d, %Y")
  full_schedule_2019[[i]]$tm_short = team_sched %>% html_nodes(".left:nth-child(4)") %>% html_text
  full_schedule_2019[[i]]$location = team_sched %>% html_nodes("#team_schedule td:nth-child(5)") %>% html_text %>% str_replace("@", "Away")
  full_schedule_2019[[i]]$location[full_schedule_2019[[i]]$location == ""] <- "Home"
  full_schedule_2019[[i]]$opp = team_sched %>% html_nodes(".left:nth-child(6)") %>% html_text
  full_schedule_2019[[i]]$runs = team_sched %>% html_nodes(".right:nth-child(8)") %>% html_text %>% as.numeric()
  full_schedule_2019[[i]]$runs_allowed = team_sched %>% html_nodes(".right:nth-child(9)") %>% html_text %>% as.numeric()
}

(full_schedule_2019 = rbindlist(full_schedule_2019))
write.csv(full_schedule_2019, "2019_schedule.csv")
```

```{r player_spending, eval = F, include = F}
# Save the html file
full_web = read_html("https://www.spotrac.com/mlb/contracts//")

# Save the data (names, position, team, age, years in league, contract amount, and avg contract amount/year) as a list

players = full_web %>% html_nodes("#main a") %>% html_text
pos = full_web %>% html_nodes(".player+ td") %>% html_text
team = full_web %>% html_nodes("td:nth-child(3)") %>% html_text %>% str_extract("\\w{2,3}")
age = full_web %>% html_nodes("td:nth-child(4)") %>% html_text %>% as.numeric()
yrs_league = full_web %>% html_nodes("td:nth-child(5)") %>% html_text %>% as.numeric()
full_contr = full_web %>% html_nodes(".small+ td.right") %>% html_text %>% str_remove_all("\\$") %>% str_remove_all(",") %>% as.numeric()
avg = full_web %>% html_nodes(".right+ td.right") %>% html_text %>% str_remove_all("\\$") %>% str_remove_all(",") %>% as.numeric()

(player_data = data.frame(players, pos, team, age, yrs_league, full_contr, avg))

write.csv(player_data, "player_contract.csv")
```

```{r old_team_spending, eval = F, include = F}
website = read_html("https://www.baseball-reference.com/leagues/MLB/2019-misc.shtml")

team = website %>% html_nodes("th a") %>% html_text
payroll = website %>% html_nodes(".right:nth-child(11)") %>% html_text %>% str_remove_all("\\$") %>% str_remove_all(",") %>% as.numeric()

team_spending = data.frame(team, payroll)
write.csv(team_spending, "team_spending2019.csv")
```

```{r Records, eval = F, include = F}
# Choose what years we want to pull
years = as.character(1995:2019)

# Set starting point
start = html_session("https://www.baseball-reference.com/leagues/MLB/")

# Create list to store information in
years_data = list()
years_standing = list()

# Manually go into each season
for(i in years) {
  # First, pull only the yearly stats per team
  # Set year data, year standings
  year_team = start %>% follow_link(i) %>% read_html()
  
  # Pull data frame
  years_data[[i]]$df = start %>% html_nodes("#teams_standard_batting tbody .right , #teams_standard_batting tbody .left") %>% html_text %>% matrix(ncol = 29, byrow = TRUE) %>% as.data.frame()
  
  # Change names of header
  names_holder = page %>% html_nodes("#teams_standard_batting .poptip") %>% html_text
  names(years_data[[i]]$df) = names_holder
  
  # Add year column
  years_data[[i]]$year = rep(i, nrow(years_data[[i]]$df))
  
  
  # Now, pull the standings for each team
  standings = start %>% follow_link(i) %>% follow_link("MLB Standings")
}



```

Hitting Stats

```{r}
# Create our list
stats_by_year_list = list()

# Choose what years to pull from
years = as.character(1901:2019)

# Start our HTML session
start = html_session("https://www.baseball-reference.com/leagues/MLB/")

# Now, we will create a loop, where for each year in baseball history
# The batting stats will be saved in the list

for(i in 1:length(years)) {
  # String of i
  b = as.character(i)
  
  # Find each year's page
  year_batting_page = start %>% follow_link(years[i]) %>% read_html()
  
  # Create our df of all the data
  stats_by_year_list[[b]] = year_batting_page %>% html_nodes("#teams_standard_batting tbody .left , #teams_standard_batting tbody .right") %>% html_text %>% matrix(ncol = 29, byrow = TRUE) %>% as.data.frame()
  
  # Change the column titles to match its meaning
  names(stats_by_year_list[[b]]) = year_batting_page %>% html_nodes("#teams_standard_batting .poptip") %>% html_text
  
  # Drop league averages, or last row
  n = nrow(stats_by_year_list[[b]])
  stats_by_year_list[[b]] = stats_by_year_list[[b]][-n,]
  
  # Create our year column, append to our data
  year = rep(years[i], n - 1)
  stats_by_year_list[[b]] = data.frame(stats_by_year_list[[b]], year)
}

for(i in years) {
  print(dim(stats_by_year_list[[as.character(i)]]))
}

yearly_batting_stats = bind_rows(stats_by_year_list, .id = "column_label")

write.csv(yearly_batting_stats, "yearly_batting_stats.csv")
```

#### Playoff teams historical

Data cleanup to include a column determining if a team made the playoffs or not

```{r, eval = F, message = F}
# Bring data in
team_data_historical = read.csv("teams.csv", na.strings = c("", " ", "NA"))
yearly_batting = read.csv("yearly_batting_stats.csv")[,-c(1, 2)]

# Look at the structure
#glimpse(team_data_historical)

# Choose columns relating to wins
win_history = team_data_historical %>% select(yearID, teamIDBR, DivWin, WCWin, LgWin, WSWin)
# Change the column names
names(win_history) = c("year", "tm_short", "DivWin", "WCWin", "LgWin", "WSWin")

# Add a new column, playoffs, if the team made playoffs
win_history = win_history %>% mutate(playoffs = ifelse(DivWin == "Y" | WCWin == "Y" | LgWin == "Y" | WSWin == "Y", T, F))
# Fix it NA's in playoff column
win_history$playoffs[is.na(win_history$playoffs)] = F

# Create a new column
playoff_history = win_history %>% select(year, tm_short, playoffs)

# Create subsets starting from 1975 to present day
subset = 1975:2020

yearly_batting_sub = yearly_batting[which(yearly_batting$year %in% subset),]
playoff_subset = win_history[which(win_history$year %in% subset),]

# Now, to join the data together
stat_playoff_full = yearly_batting_sub %>% left_join(playoff_history, by = c("Tm" = "tm_short", "year" = "year"))

(stat_playoff_full = stat_playoff_full[names(stat_playoff_full)[c(1, 30, 2:29, 31)]])

# Change the names of the columns to better reflect meaning
names(stat_playoff_full) = c("Tm", "year", "num_batters", "BatAge", "R.G", "G", "PA", "AB", "R", "H", "B2", "B3", "HR", "RBI", "SB", "CS", "BB", "SO", "BA", "OBP", "SLG", "OPS", "OPS_plus", "TB", "GDP", "HBP", "SH", "SF", "IBB", "LOB", "playoffs")

write.csv(stat_playoff_full, "clean_playoff_data.csv")
```

#### Pitching Stats

Pulling pitching stats

```{r}
# Create our list
stats_by_year_list = list()

# Choose what years to pull from
years = as.character(1901:2019)

# Now, we will create a loop, where for each year in baseball history
# The batting stats will be saved in the list

for(i in 1:length(years)) {
  # First, create our URL
  URL = paste("https://www.baseball-reference.com/leagues/MLB/", years[i], "-standard-pitching.shtml", sep = "")
  
  # String of i to store as a list with years for each name
  b = as.character(i)
  
  # Find each year's page
  year_pitching_page = read_html(URL)
  
  # Create our df of all the data
  stats_by_year_list[[b]] = year_pitching_page %>% html_nodes("#teams_standard_pitching tbody .right , #teams_standard_pitching tbody .left") %>% html_text() %>% matrix(ncol = 36, byrow = TRUE) %>% as.data.frame()
  
  # Change the column titles to match its meaning
  names(stats_by_year_list[[b]]) = year_pitching_page %>% html_nodes("#teams_standard_pitching .poptip.center") %>% html_text()
  
  # Drop league averages, or last row
  n = nrow(stats_by_year_list[[b]])
  stats_by_year_list[[b]] = stats_by_year_list[[b]][-n,]
  
  # Create our year column, append to our data
  year = rep(years[i], n - 1)
  stats_by_year_list[[b]] = data.frame(stats_by_year_list[[b]], year)
}

for(i in years) {
  print(dim(stats_by_year_list[[as.character(i)]]))
}

yearly_pitching_stats = bind_rows(stats_by_year_list, .id = "column_label")

yearly_pitching_stats

write.csv(yearly_pitching_stats, "yearly_pitching_stats.csv")
```

Cleaning a prior data frame - converting this into per game stats

```{r}
# Pull in data
stat_playoff_full = read.csv("clean_playoff_data.csv")

# Create a vector of columns that will be changed to per game
col_names = c("PA", "AB", "R", "H", "B2", "B3", "HR", "RBI", "SB", "CS", "BB", "SO", "TB", "GDP", "HBP", "SH", "SF", "IBB", "LOB")

# Create a vector of names of columns that won't be changed
untouched = setdiff(names(stat_playoff_full), col_names)

# Original list of names
original_header = names(stat_playoff_full)

# Split between two data frames, and then work
df_change = stat_playoff_full[col_names]
df_og = stat_playoff_full[untouched]

# Divide each column by the games played
df_change = df_change / df_og$G

# Append the two data frames together again
playoff_stats_pergame = cbind(df_change, df_og)

# Fix order of columns
playoff_stats_pergame = playoff_stats_pergame[original_header]

# Save as a csv again
write.csv(playoff_stats_pergame, "playoff_stats_pergame.csv")
```

#### Combining the Data

Making the full data set with both hitting and pitching stats

```{r}
# Pull in our data sets
stat_half_full = read.csv("clean_playoff_data.csv")
year_batting_stats = read.csv("yearly_pitching_stats.csv")

# Left join by stat_half_full, just to ensure that all the necessary columns are brought in
# Match by year and Tm
full_stats = stat_half_full %>% left_join(year_batting_stats, by = c("Tm", "year"), suffix = c("_own", "_allowed"))

# First, remove repeat columns and unnecessary data
del_cols = c("X", "column_label", "G_allowed", "GS", "GF")

# Find the columns that match, then remove
rem_cols = which(names(full_stats) %in% del_cols)
full_stats = full_stats[,-rem_cols]

# Time to fix the names of the columns
fix_cols = c("Tm", "year", "BatAge", "G_own", "playoffs", "X.P", "PAge", "SV", "ERA.")
repl_cols = c("Team", "Year", "BatAvgAge", "Games_Played", "Playoffs", "num_pitchers", "PitAvgAge", "Saves", "ERA_plus")

# Create an index of headers that need to be changed
fix_cols_index = which(names(full_stats) %in% fix_cols)

# Replace the columns to have the correct headers
names(full_stats)[fix_cols_index] = repl_cols

# Save our data
write.csv(full_stats, "full_data.csv")
```

Data Cleanup / Conversion to Per Game Stats, and Rankings per stat per year

```{r subset}
# Load data
full_data = read.csv("full_data.csv")[,-1]

# Store unnecessary headers as a vector, to remove later
rem_cols = c("BatAvgAge", "R_own", "PitAvgAge", "W", "L", "W.L.", "H_allowed", "HR_allowed", "BB_allowed", "SO_allowed")

# Remove columns
rem_index = which(names(full_data) %in% rem_cols)
partial_data = full_data[,-rem_index]

# Save Partial Data
write.csv(partial_data, "clean_subset.csv")
```

```{r PerGame}
# Pull in data
partial_data = read.csv("clean_subset.csv")[,-1]

# Split into 2 data frames of per-game and non per-game stats
# First, we'll save the per game variables as its own vector
per_game_cols = c("Team", "Year", "BA", "OBP", "SLG", "OPS", "OPS_plus", "R.G", "ERA", "H9", "HR9", "BB9", "SO9", "SO.W", "Playoffs")

# Now, to split the data frame into two groups
# Create an index to keep code clean
pg_index = which(names(partial_data) %in% per_game_cols)

# Split data frames
total_stats = partial_data[,-pg_index]
pg_stats = partial_data[,pg_index]

# Convert all total stats to per game stats
pg_converted = total_stats / total_stats$Games_Played

# Combine the two data frames together again
pergame_data = cbind(pg_stats, pg_converted)

# Save this new per game data
write.csv(pergame_data, "pergame_stats_fullsub.csv")
```

```{r Ranking}
# Pull in our prior per game stats
pergame_data = read.csv("pergame_stats_fullsub.csv")[,-1]

# Save all the columns that will be converted to a rank basis
# We don't need the columns regarding the Team, Year, or number of games played
rank_cols = names(pergame_data)[-which(names(pergame_data) %in% 
                                         c("Team", "Year", "Games_Played", "Playoffs"))]

# Now, create a loop to convert each column to a rank, 
# based off how each team has done comparatively for that year
for(i in 1:length(rank_cols)) {
  # First, we group by the year
  pergame_data = pergame_data %>% group_by(Year) %>% 
    # Then, we create our new column
    mutate(
           # Where we rank each row by the season's stats
           rank(!! rlang::sym(rank_cols[i]), ties.method = "average")
             )
  # Rename the last column to be the column name, with the suffix "_ranked"
  # First, we want to call the last 
  n = ncol(pergame_data)
  names(pergame_data)[n] = paste(rank_cols[i], "_ranked", sep = "")
}

# Now, remove all the original columns
drop_index = which(names(pergame_data) %in% rank_cols)
ranked_data = pergame_data[,-drop_index]


# Save our file
write.csv(ranked_data, "ranked_fullsub.csv")
```



Team Spending

```{r team_spending}
# Start on the misc page
start = html_session("https://www.baseball-reference.com/leagues/MLB/misc.shtml")

# Notice that team spending is only given from 1985 and onward, so we'll only pull each teams yearly spending from that
years = as.character(1985:2019)

# Now, specify the nodes that we will be pulling
# It will be in the format of: team, year, spending
# We can use rbind() to combine all of the rows together

df = data.frame()

for(i in 1:length(years)) {
  
  # First, we follow the link to the year, but then, because the teams in the league will vary between years, we'll pull the team data first
  start_year = start %>% follow_link(years[i]) %>% read_html()

  # Next, pull the necessary information from the given table
  Tm = start_year %>% html_nodes("th.left") %>% html_text()
  Payroll = start_year %>% html_nodes(".right:nth-child(11)") %>% html_text() %>% str_remove_all("\\$") %>% str_remove_all("\\,") %>% as.numeric()
  year = as.numeric(rep(years[i], length(Tm)))
  df_holder = data.frame(Tm, Payroll, year)
  df_holder = df_holder %>% arrange(Payroll) %>% mutate(order = 1:nrow(df_holder))
  df = rbind(df_holder, df)
}
write.csv(df, "yearly_spending.csv")
```

Creating a data frame that gives the number of teams that make the playoffs, given the year.

```{r}
# Create the years that are included in our data
years = 1975:2019

# Manually create the number of teams making the playoffs each year
playoff_teams_num = c(rep(4, 6), 8, rep(4, 12), 0, rep(8, 17), rep(10, 8))

# Create our data frame
playoff_num_df = data.frame(years, playoff_teams_num)

# Save this data frame for future use
write.csv(playoff_num_df, "playoff_num_df.csv")
```















